---
title: "Activity_Monitoring"
author: "Nithya"
date: "November 14, 2015"
output: html_document
---
&nbsp; 

&nbsp;

#Activity Monitor Data Analysis - Preamble
&nbsp; 


##Purpose
Purpose of this document is to analyze in detail the outputs of an Activity Monitor. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.
&nbsp; 

&nbsp; 

##Input Data Format
The data for this Analysis is available at the Link - [Activity Monitor Data][2]
The variables included in this dataset are:

 - **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)

 - **date**: The date on which the measurement was taken in YYYY-MM-DD format

 - **interval**: Identifier for the 5-minute interval in which measurement was taken
&nbsp; 

&nbsp; 

##Technical Details
The Alogorith and Sourcecode for the below analysis is avaibale in the github **repo - [Link to Repo][1]** . 

-- **R Markdown File** - PA1_template.Rmd

-- **Markdown File** - PA1_template.md

-- **HTML Output** - PA1_template.html

-- **Input Data** - activity.zip. This .zip file contains the dataset in a comma-separated-value (CSV) file - "activity.csv" and there are a total of 17,568 observations in this dataset.
&nbsp; 

&nbsp; 

&nbsp; 

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.width = 15, fig.height = 8)
```


```{r Funtions to Get Mean, echo=FALSE}
get_mean <- function(x, data_func, measure)
{

# Get the data filtered by interval / date

if (measure == "interval")
{
  data_filt <- subset(data_func, interval == x )  
}else
{
  data_filt <- subset(data_func, date == x )
}

# Remove the NAs
dataset <- data_filt[!is.na(data_filt$steps), 1]

# Calculate the mean
if(!is.na(mean(dataset, na.rm = TRUE)))
   {
      mean_set <- mean(dataset, na.rm = TRUE)
   }else{
      mean_set <- 0
   }

# Return the Mean
return(mean_set)

}

```

#Activity Monitor Data Analysis - Details
&nbsp; 


##Loading and preprocessing the data

```{r Getting Data into R, echo=FALSE}
  
  # Load the Libraries
  library(ggplot2)
  library(gridExtra)
  library(lattice)
  library(plyr)
  library(scales)  

  options("scipen"=100, "digits"=2)

  # URL for Activity Monitor Data
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  
  # Download the zip file
  if (!dir.exists("./activity_monitor")) {dir.create("./activity_monitor")}
  download.file(url, "./activity_monitor/activity_data.zip")
  
  # Get the file name (activity.csv) and unzip the .zip file
  fname <- unzip("./activity_monitor/activity_data.zip", list = TRUE)$Name
  unzip("./activity_monitor/activity_data.zip", files = fname, exdir = "./activity_monitor", overwrite = TRUE)
  
  # Get the location for reading the file for futher processing
  fname <- paste("./activity_monitor", fname, sep = "/")
 
```

The Data has been loaded from the URL into the loation **`r fname`**.

&nbsp; 


##Exploratory Analysis

This paper intends to identify patterns in data and answer some very specific questions, namely:

1. What is mean total number of steps taken per day?

2. What is the average daily activity pattern?

3. Are there differences in activity patterns between weekdays and weekends?

We also need to analyze the effects of keeping / removing NA values.

&nbsp; 

&nbsp;

**Plot#1 - Total number of Steps taken Per Day (keeping the NA values)**

```{r Plot 1 - steps taken per day (keeping the NA values), echo=FALSE}
  
  # Read the data and format it
  data <- read.csv(fname, stringsAsFactors = FALSE)
  
  data_raw <- data[ ,1:2]
  data_raw$steps <- as.numeric(as.vector(data_raw$steps))
  data_raw$date <- as.Date(as.vector(data_raw$date))
  
  # Aggregate the data to find total number of steps per day
  total_raw <- aggregate(data_raw$steps, by = list(data_raw$date), FUN = sum)
  names(total_raw) <- c("Date", "Steps")
  
  # Get the mean and median for the total number of steps per day
  mean_raw <- mean(total_raw$Steps, na.rm = TRUE)
  median_raw <- median(total_raw$Steps, na.rm = TRUE)
  
  # Plot the histogram and highlight the mean and median
  g0 <- ggplot(total_raw, aes(x = Date, y = Steps), legend = TRUE)
  gplot0 <- g0 + geom_histogram(stat = "identity", alpha = 1/3, fill = "blue") + 
    ylab("Total #of Steps/Day") + xlab("Date (dd-mm-yyy)") + labs(title = "Steps taken per day (with NAs in data)") + 
    scale_x_date(labels = date_format("%d-%m-%Y")) +
    theme(plot.title = element_text(size = 14, face = "bold", colour = "brown"), axis.text=element_text(size=12, angle = 90)) +
    geom_hline(aes(yintercept = mean(Steps, na.rm = TRUE), color = paste("Mean = ", round(mean(Steps, na.rm = TRUE),2))), data = total_raw, show_guide = TRUE) + 
    geom_hline(aes(yintercept = median(Steps, na.rm = TRUE), color = paste("Median = ", round(median(Steps, na.rm = TRUE),2))), data = total_raw,  show_guide = TRUE) + 
    guides(colour = guide_legend("Summary"))
  
  print(gplot0)
  
```

&nbsp; 

**Plot#2 - Total number of Steps taken Per Day (removing the NA values)**

```{r Plot 2 - steps taken per day (removing the NA values), echo=FALSE}
  
  # Find the total number of NAs
  invalid <- nrow(data[!is.na(data$steps), ])
  
  # Get the data and process it to remove the NAs. NAs are replaced by mean of total number of steps aggregated for the specific Interval
  data_valid <- data[ ,1:2]

  for(i in 1:nrow(data_valid))
      {
        if(is.na(data_valid[i, 1])) 
          {
            data_valid[i, 1] <- get_mean(data[i, 3], data, "interval") 
          }
      }
  
  data_valid$steps <- as.numeric(as.vector(data_valid$steps))
  data_valid$date <- as.Date(as.vector(data_valid$date))
  
  # Aggregate the data by calculating the total number of steps per date
  total <- aggregate(data_valid$steps, by = list(data_valid$date), FUN = sum)
  names(total) <- c("Date", "Steps")
  
  # Calculate the mean and median of total number of steps per date
  mean_total <- mean(total$Steps)
  median_total <- median(total$Steps)
  
  # Plot the histogram and highlight the mean and median
  g1 <- ggplot(total, aes(x = Date, y = Steps), legend = TRUE)
  gplot1 <- g1 + geom_histogram(stat = "identity", alpha = 1/3, fill = "blue") + 
                 ylab("Total #of Steps/Day") + xlab("Date (dd-mm-yyy)") +
                 labs(title = "Steps taken per day (removing NAs from Data)") + scale_x_date(labels = date_format("%d-%m-%Y")) +
                 theme(plot.title = element_text(size = 14, face = "bold", colour = "brown"), axis.text.x=element_text(size=12, angle = 90)) +
                 geom_hline(aes(yintercept = mean(Steps), color = paste("Mean = ", round(mean(Steps),2))), data = total, show_guide = TRUE) + 
                 geom_hline(aes(yintercept = median(Steps), color = paste("Median = ", round(median(Steps),2))), data = total,  show_guide = TRUE) + 
                 guides(colour = guide_legend("Summary"))
  
  #grid.arrange(gplot0, gplot1, nrow = 2, ncol=1)
  print(gplot1)
  
```
&nbsp; 

*Total Number of NAs replaced with appropriate values = `r invalid`*
&nbsp; 

As is evident from these graphs, There is a no variation in the mean and a very negligible variation in the median of the number of steps taken per day after removal of the NAs.

**The Mean Total Number of Steps taken per day - **

- *After Removing NA = `r round(mean_total,2)`*

- *Before Removing NA = `r round(mean_raw, 2)`*


**The Median of Total Number of Steps taken per day - **

- *After Removing NA = `r round(median_total,2)`*

- *Before Removing NA = `r round(median_raw, 2)`*

The Algorithm used to remove the NAs is fairly straightforward. NAs are replaced by the mean of total number of steps aggregated for the specific interval. 
For Example, if we encounter and NA on a date 01/10/2012 and Interval = 5, then the mean of the total number of steps taken for the interval 5 for non-NA observations is calculated and assigned here.
&nbsp; 

*Another observation is the peaking of activities around 22-23 November (reasons to be established)*
&nbsp; 

&nbsp;

&nbsp;

**Plot#3 - Average Daily Activity Pattern**

```{r Plot 3 - average daily activity pattern, echo=FALSE}

  # Get the data and process it to remove the NAs. NAs are replaced by mean of total number of steps aggregated for the specific interval
  data_int <- data[ , -2]
  
  for(i in 1:nrow(data_int))
  {
    if(is.na(data_int[i, 1])) 
    {
      data_int[i, 1] <- get_mean(data[i, 3], data, "interval")
    }
  }
  
  data_int$steps <- as.numeric(as.vector(data_int$steps))
  data_int$interval <- as.numeric(as.vector(data_int$interval))
  
  # Aggregate the data by finding the mean on total number of steps per interval
  average <- aggregate(data_int$steps, by = list(data_int$interval), FUN = mean)
  names(average) <- c("Interval", "Steps")
  
  # Get the max value of mean of total number of steps and also get the specific interval
  max_cord <- average[order(-average$Steps), ]
  max_cord <- max_cord[1, ]
  
  # Plot the line graph and highlight the maximum
  g2 <- ggplot(average, aes(x = Interval, y = Steps), legend = TRUE)
  gplot2 <- g2 + geom_line() + 
                 geom_point(data = max_cord, aes(x = Interval, y = Steps, colour = paste("Max Value: Interval = ", round(Interval,2), ", #Steps = " ,round(Steps,2))), size = 4, show_guide = TRUE) + 
                 geom_hline(aes(yintercept = mean(Steps), color = paste("Mean = ", round(mean(Steps),2))), data = average, show_guide = TRUE) + 
                 geom_smooth(se = FALSE, color="green") +
                 ylab("Avg. #of Steps/Intv.") + labs(title = "Average daily activity pattern") +
                 theme(plot.title = element_text(size = 14, face = "bold", colour = "brown")) +
                 guides(colour = guide_legend("Summary"))
  
  print(gplot2)

```
&nbsp; 

The green smoother line indicates that the activity levels start at a very low value, peak at the range of intervals 800 to 900 and then declining to another low. 
&nbsp; 

**Some generic observations that can be made about the Activity Pattern based on intervals is that - **

- *There are several peaks and crests but the most predominant one is in the range of 800 - 900 or 8.00 AM to 9.00 AM*

- *The max activity is during the 5 minute interval of 835 (8.35 AM) where the average number of steps taken is 206.17*

- *There is an overall reduction of activities post 2000 or 8.00 PM*
&nbsp; 

The Algorithm used to remove the NAs is same as for the Histograms on total number of steps taken. NAs are replaced by mean of total number of steps aggregated for the specific interval. 
For Example, if we encounter and NA on a date 01/10/2012 and Interval = 5, then the mean of the total number of steps taken for the interval 5 for non-NA observations is calculated and assigned here.
&nbsp; 

&nbsp;


**Plot#4 - Differences in Activity Patterns between Weekdays and Weekends** 

```{r Plot 4 - activity patterns between weekdays and weekends, echo=FALSE, fig.height = 10}
  
  # Get the data and process it to remove the NAs. NAs are replaced by mean of total number of steps aggregated for the specific interval
  data_wkd <- data
  
  for(i in 1:nrow(data_wkd))
  {
    if(is.na(data_wkd[i, 1])) 
    {
      data_wkd[i, 1] <- get_mean(data[i, 3], data, "interval")
    }
  }
  
  data_wkd$steps <- as.numeric(as.vector(data_wkd$steps))
  data_wkd$interval <- as.numeric(as.vector(data_wkd$interval))
  data_wkd$date <- as.Date(as.vector(data_wkd$date)) 
  
  # Divide the data into 2 groups based on whether the date is a Weekday / Weekend
  data_wkd <- mutate(data_wkd, wkd = ifelse(weekdays(data_wkd$date) %in% c("Sunday", "Saturday"), "WEEKENDS", "WEEKDAYS"))

  # Get the Aggregated values by calculating the mean of number of steps per interval
  average_wkd <- aggregate(data_wkd$steps, by = list(data_wkd$wkd, data_wkd$interval), FUN = mean)
  names(average_wkd) <- c("Wkd", "Interval", "Steps")
  
  # Get the Averages - for weekends and weekdays separately
  mean_weekends <- mean(data_wkd[data_wkd$wkd == "WEEKENDS", "steps"], na.rm = TRUE)
  mean_weekdays <- mean(data_wkd[data_wkd$wkd == "WEEKDAYS", "steps"], na.rm = TRUE)
  df <- data.frame(Wkd = c("WEEKDAYS", "WEEKENDS"), mean = c(mean_weekdays, mean_weekends))
  
  # Plot the line graph with 2 facets - Weekend and Weekday and also highlight the respective means
  g3 <- ggplot(average_wkd, aes(x = Interval, y = Steps), legend = TRUE)
  gplot3 <- g3 + geom_line() + 
                 facet_wrap(~Wkd, ncol = 1) +
                 ylab("Avg. #of Steps/Intv.") + labs(title = "Activity patterns weekdays Vs weekends") +
                 theme(plot.title = element_text(size = 14, face = "bold", colour = "brown")) +
                 geom_hline(data = df, aes(yintercept = mean, color = paste(Wkd, " = ", round(mean,2))), show_guide = TRUE) + 
                 geom_smooth(se = FALSE, color="green") +
                 guides(colour = guide_legend("Mean"))
                 
  
  #grid.arrange(gplot2, gplot3, nrow = 2, ncol=1)
  print(gplot3)

    
```  
&nbsp;

**There are some important variations to be noted in the activity patterns between weekdays and weekends:**

- *The mean activity over the weekdays (`r mean_weekdays`) is lesser than the mean activity during the weekends (`r mean_weekends`)*

- *There is a stark peak of activities over the weekdays between 800 to 900 or in other words 8.00 AM to 9.00 AM*

- *Referring to the green smoothers, The peak activity on the weekdays is in the morning time, whereas the peak over the weekends shifts more towards the later part*

- *Quite obviously the activities reduce in the later part of the day. But on weekends this reduction in activities is post 2000 or 8.00 PM whereas on weekdays it is a bit earlier at approax 1800 or 6.00 PM*

&nbsp; 

&nbsp; 


#Clustering the activities
&nbsp; 

Let us try to make use of the Principle Component Analysis technique to identify any sort of clustering or grouping that can be established.
&nbsp; 

```{r Clustering Analysis - PCA}
  
  library(cluster)
  library(xtable)
  library(knitr)
  
  set.seed(12345)
  
  fit <- kmeans(average, 5, nstart = 100)
  clusplot(average, fit$cluster, color = TRUE, shade = TRUE, main = "Clustering the Number of steps using PCA")
```
```{r}
  plot(average$Interval, average$Steps, pch = 19, col = fit$cluster, xlab = "Interval", ylab = "Average numer of Steps", main = "Activities per Cluster")
```
&nbsp; 

We have divided the Activity Levels into 5 Clusters. Below are the Cluster Centers (Means)
```{r}
  
  cent <- fit$centers
  names(cent) <- c("Interval", "Average Activity (Mean number of Steps)")
  cent <- cent[sort.list(cent[,2],decreasing = TRUE ), ]
  kable(cent, digits = 2,align = "c", row.names = FALSE)
  
```

&nbsp; 

We can see that the highest activity intervals are centered around 8.00 AM (approx) followed by intervals centered around 5.00 PM (approx)
The Lowest activity interval is centered around 2.00 AM (approx).

&nbsp; 

&nbsp; 


[1]: <https://github.com/nithsubr/RepData_PeerAssessment1> "Link to Repo" 
[2]: <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip> "Activity Monitor Data"